Concurrent Programming with CXL

### 2020-06-19 book/paper.tex
Improved syntax

It is often the case that methods that update shared variables through
pointers do not actually have to return a result to be useful.
In CXL, if you want to invoke a method that does not return a result
or if you are simply not interested in it,
you have to use a \texttt{call} statement like so:
\begin{code}
\begin{verbatim}
call m a;
\end{verbatim}
\end{code}

@@

While the CVM does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{Verbatim}
\end{code}

@@

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            call stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
    if left < right:
        call lock &(forks[left]);
        call lock &(forks[right]);
    else:
        call unlock &(forks[right]);
        call unlock &(forks[left]);
    ;
\end{Verbatim}
\end{code}

@@

CCSs are easy to implement using busy waiting:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
let blocked = True:
    while blocked:
        call lock();
        if application-condition-holds:
            blocked = False;
        else:
            call unlock():
@cr: application-code;
call unlock():
\end{Verbatim}
\end{code}

@@

\begin{code} 
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def wait(c):
    call V &(mutex);
    call P c;
;
def signal(c):
    call V c;
    call P &(mutex);
;
\end{Verbatim} 
\end{code}

@@

def Condition(lk):
    result = dict{ .lock: lk, .waiters: bagEmpty() };
;
def wait(c):
    let lk = (^c).lock, blocked = True, cnt = bagCount((^c).waiters, nametag()):
        call bagAdd(&((^c).waiters), nametag());
        ^lk = False;
        while blocked:
            atomic:
                if (not (^lk)) and (bagCount((^c).waiters, nametag()) <= cnt):
                    ^lk = True;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def notify(c):
    let waiters = (^c).waiters:
        if waiters != bagEmpty():
            call bagRemove(&((^c).waiters), bagChoose(waiters));
        ;
    ;
;
def notifyAll(c):
    (^c).waiters = bagEmpty();
;
\end{Verbatim}
\end{code}
\caption{Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}

@@

\begin{code}
\begin{Verbatim}
while forks[left]:
    call wait &(conds[left]);
;
while forks[right]:
    call wait &(conds[right]);
;
\end{Verbatim}
\end{code}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del lv} & delete\\
\hline
\texttt{call e} & evaluate \texttt{e}\\
\hline
\texttt{assert b [, x]} & \texttt{b} is a Boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
\hline
\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
\hline
\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
\hline
\texttt{while b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
\hline
\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
\hline
\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}

>>>>

@@

While the CVM does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
atomic_inc &(count);
\end{Verbatim}
\end{code}

@@

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
    if left < right:
        lock(&forks[left]);
        lock(&(orks[right]);
    else:
        unlock(&forks[right]);
        unlock(&forks[left]);
    ;
\end{Verbatim}
\end{code}

@@

CCSs are easy to implement using busy waiting:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
let blocked = True:
    while blocked:
        lock();
        if application-condition-holds:
            blocked = False;
        else:
            unlock():
@cr: application-code;
unlock():
\end{Verbatim}
\end{code}

@@

\begin{code} 
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def wait(c):
    V(&mutex);
    P(c);
;
def signal(c):
    V(c);
    P(&mutex);
;
\end{Verbatim} 
\end{code}

@@

def Condition(lk):
    result = dict{ .lock: lk, .waiters: bagEmpty() };
;
def wait(c):
    let lk = (^c).lock, blocked = True, cnt = bagCount((^c).waiters, nametag()):
        bagAdd(&(^c).waiters, nametag());
        ^lk = False;
        while blocked:
            atomic:
                if (not (^lk)) and (bagCount((^c).waiters, nametag()) <= cnt):
                    ^lk = True;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def notify(c):
    let waiters = (^c).waiters:
        if waiters != bagEmpty():
            bagRemove(&(^c).waiters, bagChoose(waiters));
        ;
    ;
;
def notifyAll(c):
    (^c).waiters = bagEmpty();
;
\end{Verbatim}
\end{code}
\caption{Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}

@@

\begin{code}
\begin{Verbatim}
while forks[left]:
    wait(&conds[left]);
;
while forks[right]:
    wait(&conds[right]);
;
\end{Verbatim}
\end{code}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del lv} & delete\\
\hline
\texttt{assert b [, x]} & \texttt{b} is a Boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
\hline
\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
\hline
\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
\hline
\texttt{while b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
\hline
\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
\hline
\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}
